Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-08-30 manuscript/manuscript_file.Rmd
Revised Figure 6 and text

## Emission Rates[a]

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

@@

Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area. The largest lake size class (\> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area. This lake size class had the greatest N~2~O flux, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies. Despite widespread N~2~O undersaturation, all but two ecoregions were a net source of N~2~O to the atmosphere **[(Table 1)]**. Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface **[(Table 1)]**. The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

While total waterbody surface area contributed to differences in N~2~O flux among ecoregions, large cumulative surface area does not necessarily equate to large fluxes. The Upper Midwest ecoregion had the second largest waterbody surface area, but only the sixth largest flux. This is due to a mean saturation (mean = 1, 95% credible interview: 0.99 - 1.01) indicative of equilibirum with the atmosphere, which translates to very low fluxes.

At the national scale, we estimate that CONUS waterbodies emit `r table1_dat %>% janitor::clean_names() %>% filter(spatial_domain == "national") %>% select(n2o_flux) %>% pull()` metric tons N~2~O year^-1^. While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ [@epa_2023]. Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC [@epa_2023], with the balance emanating from streams and rivers. Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00 metric tons N~2~O year^-1^), stationary combustion (84,000 metric tons N~2~O year^-1^), and manure management (66,000 metric tons N~2~O year^-1^) [@epa_2023].

@@

![Figure 6: N~2~O flux by ecoregion and lake size category. Mean (points) and 95% credible interval (horizontal bars) reflect the distribution of 500 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxBySizeEco.tiff)

@@

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include measurements at the nations largest lakes and reservoirs.

>>>>

## Emission Rates

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **[(Fig. 6A)]**, which is consistent with the relationship between N~2~O saturation and waterbody size **[(Fig. 5A,B)]**. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

![Figure 6: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. C) N~2~O flux plotted against the five lake size categories used in the survey design. Across all plots, points and bars reflect the mean (points) and 95% credible interval of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Total N~2~O flux from the surface of CONUS waterbodies is a function of N~2~O emission rates and waterbody surface areas. The four smallest size classes (\< 50 Ha) in the survey design contained `r round(proportionSmallNumber, 2) * 100`% of all lakes in the population, but contributed only `r round((FluxBySize %>% filter(size_cat != "50_max") %>% select(estimate) %>% summarise(tot_small = sum(estimate)) %>% pull())/(FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull()) * 100, 0)`% to the N~2~O flux from CONUS waterbodies **[(Fig. 6C)]**. The reason for this pattern is two-fold. First, lakes in the smaller size classes constituted only `r round(proportionSmallSA, 2) * 100`% of total waterbody surface area, thereby limiting their N~2~O flux. Secondly, widespread N~2~O uptake in these size classes largely offset the very high emission rates predicted to occur in a minority of the population **[(Fig. 6A)].** Waterbodies in the largest size class (\> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area. This lake size class had the greatest N~2~O flux (mean = `r FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull() %>% round(0)` metric tons N~2~O year^{-1}), but the flux estimate is poorly constrained with a 95% credible interval of `r FluxBySize %>% filter(size_cat == "50_max") %>% select(LCL) %>% pull()` - `r FluxBySize %>% filter(size_cat == "50_max") %>% select(UCL) %>% pull()` metric tons N~2~O year^{-1} **[(Fig. 6C)]**. While uncertainty in N~2~O emission rates was smallest for large lakes **(Fig. 6A)**, these small uncertainties propagated to large uncertainties in flux when expressed across the surface area of large waterbodies.

Despite widespread N~2~O undersaturation, only two ecoregions had flux estimates less than zero (Xeric and Coastal Plains), but the credible intervals for those ecoregions were broad and include zero **(Table 1)**. Only the Temperate Plains, Southern Plains, and Southern Appalachians had flux estimate credible intervals that did not overlap with zero and these ecoregions functioned as net N~2~O sources.  Overall, flux estimates overlapped across all nine ecoregions, which is somewhat surprising given the large differences in environmental conditions across the ecoregions.

At the national scale, we estimate that CONUS waterbodies are a source of N~2~O, emitting `r table1_dat %>% janitor::clean_names() %>% filter(spatial_domain == "national") %>% select(n2o_flux) %>% pull()` metric tons N~2~O year^-1^; however, this estimate is not well constrained and the 95% credible interval includes zero, thus we cannot conclude with statistical confidence whether CONUS waterbodies are a net source of sink of N~2~O. While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ [@epa_2023]. Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC [@epa_2023], with the balance emanating from streams and rivers. Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00 metric tons N~2~O year^-1^), stationary combustion (84,000 metric tons N~2~O year^-1^), and manure management (66,000 metric tons N~2~O year^-1^) [@epa_2023].  Whether CONUS waterbodies are a net source or sink of N~2~O, their contribution to the US N~2~O budget is small.

@@

@@

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nations largest lakes including an assesment of intra-waterbody spatial variation in N~2~O, which can be important [@Salk_etal_2019, @Borges_etal_2023], as well as careful accounting of measurement an analytical uncertainty which can propogate to large uncertainties in flux estimates.
